Collective excitations in the 2- and 3-dimensional extended Hubbard model

### 2023-11-24 paper/main.tex
tex und cluster

The Hubbard model has been studied in a plethora of previous studies. 
Early studies proved the existence of eigenstates to the Hubbard Hamiltonian that exhibit off-diagonal long-range order,
encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact $SO(4)$ symmetry was discovered,
which allows a degeneracy of superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence can be argued by the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model exactly onto the repulsive one, exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The previously mentioned SC and CDW phases map to different spin operators \cite{zitko15,lieb89}.
\newline
Many recent experimental and theoretical studies examined the driven systems in the hope of inducing superconductivity 
\cite{Nicoletti14,Krull14,Moor14,Casandruc15,patel16,sentef17,Buenemann17}.
Other studies, focusing on equilibrium systems, 
investigated the phases and various quantities therein of the said model including various additional interactions with and without doping 
\cite{Micnas88,Micnas88b,Micnas89,Dzierzawa92,Kostyrko92,Eriksson95,Staudt00,Onari04,Toschi05,Brackett16,Paki19,romer20,Sushchyev22}.
\newline
In this paper, we will restrict ourselves to the half-filled Hubbard model, including an additional intersite interaction, on a square and a simple cubic lattice.
\textcolor{red}{Für den 3D Fall finde ich keine Phasendiagramme des extended Hubbard models, ggf nochmal suchen...}
The 2D case already exhibits a wide range of possible phases, including CDW, AFM as well as $s$- and $d_{x^2 - y^2}$-wave superconductivity
\cite{Micnas88b,Tsuchiura95,Su01,Su04,ha11,Huang13,Jiang22}.
\newline
In this article, we will first employ a mean-field approximation to the interaction terms 
and then use the methodology from the so-called iterated equations of motion approach,
which has already seen success in the handling of interaction quenches \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
Here, one commutes operators from a suitable operator basis with the Hamiltonian and adds the newly occurring terms to it.
Naturally, a truncation is necessary for most practical applications, however, the approximation becomes better the more terms are included within the basis.
The functionality of this method has been compared to the density matrix formalism and has proved to be considerably more accurate \cite{Kalthoff17}.
Additionally, we will show an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Leggett) and amplitude (Higgs) modes \cite{Cea14,Krull16,Schwarz20,Fan22}.
\newline
The remainder of this paper is organized as follows:
In \autoref{sec:model} we will introduce the model and its Hamiltonian as well as the methods used in this article.
Next, we show our results in \autoref{sec:results}.
Lastly, in \autoref{sec:conclusion}, we will discuss our results.
In the appendix \textcolor{red}{(or in the supplemental material?)}, we will derive and explain our methods in detail.

\section{Model}\label{sec:model}

@@

\begin{equation}
    \begin{aligned}
        n_{k\sigma} &:= c_{\vk\sigma}^\dagger c_{\vk\sigma}\,,      &f_k     &:= c_{-\vk\down} c_{\vk\up}\,, \\
        g_{k\sigma} &:= c_{\vk\sigma}^\dagger c_{\vk+\vQ\sigma}\,,  &\eta_k  &:= c_{-\vk-\vQ\down} c_{\vk\up}\,,
    \end{aligned}
\end{equation}

@@

Now, the entire Hamiltonian and thus all expectation values only depend on
\begin{equation}
    \hat{\gamma}(\vk) := \sum_{\alpha=1}^D \cos(k_\alpha)\,,
\end{equation}
i.e. for any operator $\hat{O}$ the following relation holds 
\begin{equation}
    \langle \hat{O}_{\vk} \rangle = \langle \hat{O}_{\vk'} \rangle =: \langle \hat{O}_{\gamma} \rangle\,,
\end{equation}
if $\hat{\gamma}(\vk) = \hat{\gamma}(\vk')$.
This allows us to replace the momentum sums with energy integrals using the bare density of states
\begin{equation}
    \rho(\gamma) := \frac{1}{N} \sum_{\vk} \delta \left(\gamma - \hat{\gamma} (\vk) \right)\,.
\end{equation}
As an example, consider 
\begin{equation}
    \Delta_\text{SC} = \frac{U}{2N} \int \mathrm{d}\gamma \rho(\gamma) \langle f_{\gamma} \rangle\,.
\end{equation}
We will be using the exact densities of states for the square and the simple cubic lattice found in Ref. \cite{Hanisch97}.
Solving the mean-field equations iteratively, one obtains the groundstate phase diagram at temperature $T=0$ as shown in \autoref{fig:phase_diagram}.
Note, that the $d_{x^2 - y^2}$-superconducting state, which has been confirmed for the square lattice \cite{Micnas88b,Huang13}, cannot be described by our method. 
This is due to our restriction to terms that are proportional to $\hat{\gamma}(\vk)$.
\newline
The CDW-AFM boundary is located at $U = ZV$, where $Z$ is the coordination.
This holds for both the square lattice ($Z=4$) and the simple cubic lattice ($Z=6$) and can be seen by comparing \eqref{eqn:delta_cdw} and \eqref{eqn:delta_afm}:
Crossing the aforementioned line, changes which one of the two prefactors is larger.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Methods                                                  %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{equation}
    \begin{aligned}
        \ddt a(t) = \sum_j \ddt c_j(t) A_j &= i \sum_j c_j(t) [H, A_j] \\
        \Rightarrow \sum_j \underbrace{(A_i | A_j)}_{:=\mathcal{N_{ij}}} \ddt c_j(t) &= i \sum_j \underbrace{(A_i | [H, A_j])}_{:=\mathcal{M_{ij}}} c_j(t) \\
        \Rightarrow \mathcal{N} \ddt \vec{c}(t) &= i \mathcal{M} \vec{c}(t)\,.
    \end{aligned}
\end{equation}

@@

where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Nevertheless, we will compute $\mathcal{M}$ using the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behaviour rather than the one particle dynamics given by the mean-field approximation.

@@

>>>>

The Hubbard model has been studied in a plethora of previous studies. 
Early studies proved the existence of eigenstates to the Hubbard Hamiltonian that exhibit off-diagonal long-range order,
encouraging the model's usage for the description of high-temperature superconductivity \cite{yang89}.
Shortly afterward, an exact $SO(4)$ symmetry was discovered,
which allows a degeneracy of superconductivity (SC) and a charge-density wave (CDW) governed by an attractive on-site interaction \cite{yang90}.
This coexistence can be argued by the existence of a particle-hole transformation on bipartite lattices, 
that maps the attractive Hubbard model exactly onto the repulsive one, exhibiting antiferromagnetism (AFM) \cite{Hirsch85}.
The previously mentioned SC and CDW phases map to different spin operators \cite{zitko15,lieb89}.
\newline
Many recent experimental and theoretical studies examined the driven systems in the hope of inducing superconductivity 
\cite{Nicoletti14,Krull14,Moor14,Casandruc15,patel16,sentef17,Buenemann17}.
Other studies, focusing on equilibrium systems, 
investigated the phases and various quantities therein of the said model including various additional interactions with and without doping 
\cite{Micnas88,Micnas88b,Micnas89,Dzierzawa92,Kostyrko92,Eriksson95,Staudt00,Onari04,Toschi05,Brackett16,Paki19,romer20,Sushchyev22}.
\newline
In this paper, we will restrict ourselves to the half-filled Hubbard model, including an additional intersite interaction, on a square and a simple cubic lattice.
\textcolor{red}{Für den 3D Fall finde ich keine Phasendiagramme des extended Hubbard models, ggf nochmal suchen...}
The 2D case already exhibits a wide range of possible phases, including CDW, AFM as well as $s$- and $d_{x^2 - y^2}$-wave superconductivity
\cite{Micnas88b,Tsuchiura95,Su01,Su04,ha11,Huang13,Jiang22}.
\newline
In this article, we will first employ a mean-field approximation to the interaction terms 
and then use the methodology from the so-called iterated equations of motion approach,
which has already seen success in the handling of interaction quenches \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
Here, one commutes operators from a suitable operator basis with the Hamiltonian and adds the newly occurring terms to it.
Naturally, a truncation is necessary for most practical applications, however, the approximation becomes better the more terms are included within the basis.
The functionality of this method has been compared to the density matrix formalism and has proved to be considerably more accurate \cite{Kalthoff17}.
Additionally, we will show an explicit way to obtain various Green's functions from the aforementioned methodology.
By extension, we also obtain the spectral functions of the investigated systems and discuss the excitations found therein.
The prominent examples in the SC phase are the well-known phase (Leggett) and amplitude (Higgs) modes \cite{Cea14,Krull16,Schwarz20,Fan22}.
\newline
The remainder of this paper is organized as follows:
In \autoref{sec:model} we will introduce the model and its Hamiltonian as well our mean-field theory.
A brief overview over the iterated equations of motion approach is presented in \autoref{sec:ieom}.
Next, we show our results in \autoref{sec:results}.
Lastly, in \autoref{sec:conclusion}, we will discuss our results.
In the appendix \textcolor{red}{(or in the supplemental material?)}, we will derive and explain our methods in detail.

\section{Model and mean-field theory}\label{sec:model}

@@

\begin{equation}
    \label{eqn:operators}
    \begin{aligned}
        n_{k\sigma} &:= c_{\vk\sigma}^\dagger c_{\vk\sigma}\,,      &f_k     &:= c_{-\vk\down} c_{\vk\up}\,, \\
        g_{k\sigma} &:= c_{\vk\sigma}^\dagger c_{\vk+\vQ\sigma}\,,  &\eta_k  &:= c_{-\vk-\vQ\down} c_{\vk\up}\,,
    \end{aligned}
\end{equation}

@@

Now, the entire Hamiltonian and thus all expectation values only depend on
\begin{equation}
    \hat{\gamma}(\vk) := \sum_{\alpha=1}^D \cos(k_\alpha)\,,
\end{equation}
i.e. for any operator $O$ the following relation holds 
\begin{equation}
    \label{eqn:equal_expecs}
    \langle O_{\vk} \rangle = \langle O_{\vk'} \rangle =: \langle O_{\gamma} \rangle\,,
\end{equation}
if $\hat{\gamma}(\vk) = \hat{\gamma}(\vk')$.
This allows us to replace the momentum sums with energy integrals using the bare density of states
\begin{equation}
    \rho(\gamma) := \frac{1}{N} \sum_{\vk} \delta \left(\gamma - \hat{\gamma} (\vk) \right)\,.
\end{equation}
As an example, consider 
\begin{equation}
    \Delta_\text{SC} = \frac{U}{2N} \int \mathrm{d}\gamma \rho(\gamma) \langle f_{\gamma} \rangle\,.
\end{equation}
We will be using the exact densities of states for the square and the simple cubic lattice found in Ref. \cite{Hanisch97}.
Solving the mean-field equations iteratively, one obtains the groundstate phase diagram at temperature $T=0$ as shown in \autoref{fig:phase_diagram}.
Note, that the $d_{x^2 - y^2}$-superconducting state, which has been confirmed for the square lattice \cite{Micnas88b,Huang13}, cannot be described by our method. 
This is due to our restriction to terms that are proportional to $\hat{\gamma}(\vk)$.
\newline
The CDW-AFM boundary is located at $U = ZV$, where $Z$ is the coordination.
This holds for both the square lattice ($Z=4$) and the simple cubic lattice ($Z=6$) and can be seen by comparing \eqref{eqn:delta_cdw} and \eqref{eqn:delta_afm}:
Crossing the aforementioned line, changes which one of the two prefactors is larger.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                  Methods                                                  %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Iterated equations of motion and Green's functions}\label{sec:ieom}

@@

\begin{equation}
    \label{eqn:heisenberg}
    \begin{aligned}
        \ddt a(t) = \sum_j \ddt c_j(t) A_j &= i \sum_j c_j(t) [H, A_j] \\
        \Rightarrow \sum_j \underbrace{(A_i | A_j)}_{:=\mathcal{N_{ij}}} \ddt c_j(t) &= i \sum_j \underbrace{(A_i | [H, A_j])}_{:=\mathcal{M_{ij}}} c_j(t) \\
        \Rightarrow \mathcal{N} \ddt \vec{c}(t) &= i \mathcal{M} \vec{c}(t)\,.
    \end{aligned}
\end{equation}

@@

where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Nevertheless, we will compute $\mathcal{M}$ using the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behaviour rather than the one particle dynamics given by the mean-field approximation.
\newline
This choice enables us to write down matrix-valued Fourier-transformed Green's functions in terms of the aforementioned matrices

\begin{equation}
    \label{eqn:green_function}
    \mathcal{G}(z = \omega + i0^+) = \mathcal{N} \frac{1}{-z \mathcal{N} - \mathcal{M}} \mathcal{N} := -\mathcal{N} \mathcal{R}(z) \mathcal{N}\,,
\end{equation}

where we defined the resolvent

\begin{equation}
    \label{eqn:resolvent}
    \mathcal{R}(z) := \frac{1}{z \mathcal{N} + \mathcal{M}}\,.
\end{equation}

Here, the different matrix elements correspond to various Green's functions, depending on the operator basis.
For example, assume the operator in the basis is $f_{\vk}$, then the top left element of $\mathcal{G}$ is the Green's function

\begin{equation}
    \mathcal{G}_{00}(z = \omega +i0^+) =  G(z) = -i \int_0^{\infty} \langle [f_{\vk}(t), f_{\vk}^\dagger(0)] \rangle e^{izt} \mathrm{d}t\,.
\end{equation}

For a detailed derivation of this statement, we refer the reader to the appendix \textcolor{red}{!REF!}.

Our operator basis consists of operators of the type

\begin{equation}
    A_\gamma := \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \hat{\gamma}( \vk )) A_{\vk}\,,
\end{equation}

where $A_{\vk}$ represents each type of operator in \eqref{eqn:operators} and their Hermitian conjugates.
Different indices are taken as additional basis operators, e.g. $n_{\vk \up}$ and $n_{\vk \down}$ are taken into the basis separately.
We then order these operators in a way reminiscent of the $x$- and $p$-terms in the harmonic oscillator, i.e.

\begin{equation}
    X_i := A_i + A_i^\dagger\,,\quad P_i := A_i - A_i^\dagger\,.
\end{equation}

Our basis is then $\mathcal{B}_{XP} := {X_i, P_i}$.
If some operators would be 0 ($P_i$ for any $n$-type term) or duplicates (certain $X_i$ for $g$-type terms), we omit them.
\newline
Due to this choice of basis, the matrices occurring in \eqref{eqn:heisenberg} have a special block structure

\begin{equation}
    \mathcal{M} = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mathcal{N} = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix}\,,
\end{equation}

where $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$.
Furthermore, if all occurring expectation values are real, both matrices have large empty blocks.
This condition is fulfilled for all cases investigated in this article.
Additionally, we can utilize this block structure to rewrite \eqref{eqn:resolvent} to

\begin{equation}
    \mathcal{R}(z) \vert_X = - \check{N}^{-1/2} \frac{1}{z^2 - \underbrace{\check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}}_{:= \check{M}}} \check{N}^{-1/2}\,,
\end{equation}

where $\vert_X$ denotes the block corresponding to the $X$-terms in the basis and $\check{N} := \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger$.
An analogous form can be found for the $P$-block.

Now, the task at hand is finding the inverse $1/(z^2 - \check{M})$ for all $z$.
This can be achieved using a Lanczos tridiagonalization and subsequently a continued fraction expansion in the form of

\begin{widetext}
\begin{equation}
    \left( z^2 - \begin{pmatrix}
        z - a_0 & b_1 & 0 & 0 & \cdots \\
        b_1 & z - a_1 & b_2 & 0 & \cdots \\
        0 & b_2 & z - a_2 & b_3 & \cdots \\
        \vdots & \vdots & \vdots & \vdots & \ddots
    \end{pmatrix} \right)_{00}^{-1} = \dfrac{1}{z^2 - a_0 - \dfrac{b_1^2}{z^2 - a_1 - \dfrac{b_2^2}{ z^2 - a_2 - \hdots}}}\,\,
\end{equation}
\end{widetext}

where $a_i$ and $b_i$ are the Lanczos coefficients \cite{PettiforRecursion,ViswanathRecursion}.
In the single-band case, the coefficients approach the limit

\begin{equation}
    \label{eqn:inf_lanczos}
    a_\infty = \frac{\omega_+ + \omega_-}{2}\text{ and } b_\infty = \frac{\omega_+ - \omega_-}{4}\,,
\end{equation}

where $\omega_\pm$ represent the upper and lower band edge, respectively.
This behaviour allows us to terminate the continued fraction, if the convergence is good enough, using the so-called square root terminator

\begin{equation}
    T(\omega) = \frac{1}{2b_\infty^2} \left( \omega - a_\infty \mp \sqrt{(\omega - a_\infty)^2 - 4 b_\infty^2} \right)\,,
\end{equation}

where the negative sign is to be chosen for $\omega - a_\infty > 2b_\infty$ and $|\omega - a_\infty| < 2b_\infty$ and the positive one otherwise.

@@

\subsection{SC-CDW transition}

\subsection{CDW-AFM transition}
